Extracting sunbeams from cucumbers: Computational reproducibility is essential for archaeological science

### 2025-02-05 paper.qmd
writing conclusion

At the time of writing (January 2025) I have completed 47 reproducibility reviews of 25 manuscripts submitted to JAS. Of these, 11 have been published in JAS to date. Seven of these eleven papers fully passed the reproducibility review, resulting in a success rate, by one measure, of 63%. Four of the seven papers could be fully reproduced on my first attempt, the others required additional input from the authors. For comparison with reproducibility studies in other fields reported above, the seven fully reproducible papers should be divided by the 25 reviewed for reproducibility, resulting in a 28% success rate. Expanding the denominator to include the total number of research articles published in JAS from May 2024 (when the first article to pass the reproducibility review, @herskindComputationalLinguisticMethodology2024, was published) to January 2025 (n = 97), which are articles that could have been eligible for reproducibility review had the authors used an open source programming language (e.g. instead of commercial software such as Microsoft Excel or SPSS, etc.) the success rate is 7%. In any case, the computational reproducibility of archaeological research is generally on the low end of the distribution of values obtained from a variety of hard and soft sciences.

@@

Despite the small number of reproducibility reviews reported on here, there are patterns of common issues that point to a small set of simple tasks that have high potential to increase reproducibility. The problem of incomplete materials can be tackled in several basic ways. First, authors should use a clear and logical folder structure for their code and data. There are many simple templates to guide authors on this, e.g. @doi:10.1080/00031305.2017.1375986, @figueiredoSimpleKitUse2022, @Cookiecutter, @cooperGuideReproducibleCode2017 and @wilsonGoodEnoughPractices2017 Second, for analyses that are not highly time-consuming, authors should re-run their code more than once, and ideally on more than one machine (e.g. by another co-author of the paper), before submission to confirm everything works as expected. This would likely also prevent issues relating to path and function errors. Complex and time-consuming analyses should use pipeline or workflow management tools, e.g. GNU Make, Luigi, Snakemake, or Targets, to document the relationship of the files and folders in a machine-readable format and simplify running and re-running code [@wrattenReproducibleScalableShareable2021; @landau2021targets]. Third, authors should include in their compendium a README document that describes to a reader the folders and files contained in the project. The README file is typically the first file that a reader will look at in a compendium so it should include brief instructions to guide the user to a successful reproduction of the original results (e.g. what order to run the code files in). Fourth, authors should ensure clear, direct and obvious connections between their code and the results they present in their paper. A simple way to do this is to have one code file for each figure and table, and name the code files with the figure or table number and some key words in the caption. Another way some authors are accomplishing this is by using literate programming tools, such as Quarto and Jupyter notebooks [@quarto; @kluyver2016jupyter] that enable the research narrative and code for data analysis to be woven together in one document. Quarto was a popular tool among the JAS papers in the reproducibility review sample, for example, @vernonMethodDefiningDispersed2024 and @ragnoSheepGoatsTaxonomic2024 wrote their entire manuscript using Quarto. These four simple steps have high potential to substantially increase the computational reproducibility of archaeological science.

Our finding that dependencies are a common cause of irreproducible results is consistent with previous studies that have identified this as a widespread weakness in communicating computationally intensive research [@trisovicLargescaleStudyResearch2022; @samuelComputationalReproducibilityJupyter2024]. In our sample, issues relating to dependencies are strongly associated with the use of Python. One possible reason for this is that relative to R, Python uses more package managers, more environments, and deeper dependency chains with more complex inter-dependencies that change more rapidly [@decan2016topology; @decanEmpiricalComparisonDependency2019; @korkmazModelingImpactPython2020]. Another reason may be that there is a bigger and more established community of R users in archaeology [@schmidtToolDrivenRevolutionsArchaeological2020; @batistOpenArchaeologyOpen2024] that highly values code that is easy for others to reuse and has evolved practices to effectively communicate dependencies [e.g. @bilottiPointPatternAnalysis2024; @willExploringUtilityUnretouched2025]. The simplest way for archaeologists, to improve here is to write the names and version numbers of the software and packages they used in their README file, as we see in @herskindComputationalLinguisticMethodology2024 and @monnaStudyingSerialityMaterial2024. For more complex research projects, i.e. those using five or more packages or machine learning algorithms, authors should use dependency management tools to keep track of the packages and version numbers needed to reproduce their results. This is an active area of development, and while there are many tools currently available, the most robust and widely used include renv for R [@renv, see examples in @vernonMethodDefiningDispersed2024 and @ragnoSheepGoatsTaxonomic2024] and conda and poetry for Python [@conda2023; @poetry2023]. A more comprehensive solution, and the leading best practice for managing dependencies in many computationally intensive fields using Python in particular, is to include a Dockerfile in the compendium [@moreauContainersComputationalReproducibility2023]. This is a set of a machine- and human-readable instructions that enables a user to recreate the author's computational environment (including those requirements beyond the R or Python packages) on another computer [@nustTenSimpleRules2020]. Dockerfiles are gradually being adopted by archaeologists, see @cremaModellingDiffusionInnovation2024 and @liaoApplyingMaskRCNN2024 for examples, and most of my reproducibility reviews include a recommendation that the author include a Dockerfile to manage dependencies efficiently.

@@

@@

While not yet the norm in archaeology, the systemic verification of the computational reproducibility of the results of conditionally accepted papers is now required in many leading academic journals including the American Economic Review, American Economic Journal, Econometrica, Review of Economic Studies, Economic Journal, Journal of the American Statistical Association, American Journal of Political Science, and the Journal of Politics.

Computationally reproducible research can still be plagued by various types of errors, ranging from honest mistakes (e.g., typos in the code) to outright fraud (e.g., data fabrication). A practical hurdle in ascertaining computational reproducibility for the published literature is that the verification is resource-intensive

@@

Computational reproducibilty has become central to archaeology's status a science, especially as the definition and practice of science shifts over time. Computers have become an essential field and laboratory instrument for almost all of us, so we should document how we change our data as it flows through silicon just as carefully as we document the operating parameters of a mass spectrometer or other laboratory instrument. In making the case for reproducibilty as a core value in archaeological science

In recent years the Journal of Archaeological Science has been one of a number of archaeology journals that introduced an unusual new kind of peer review, one that evaluates the computational reproducibilty of the research submitted for publication. This journal was among the first of these and is currently by far the most active in conducting these reviews. In this article I reflect on how this new kind of review came about, why it is important for the future of archaeological science, and offer suggestions for prospective authors based on my experience as a reproducibilty reviewer for this journal on how we can ensure archaeological science remains worthy of the name.

My motivation is the view that It is customary in significant round-number anniversary -->
```

>>>>

At the time of writing (January 2025) we have completed 47 reproducibility reviews of 25 manuscripts submitted to JAS. Of these, 11 have been published in JAS to date. Seven of these eleven papers fully passed the reproducibility review, resulting in a success rate, by one measure, of 63%. Four of the seven papers could be fully reproduced on my first attempt, the others required additional input from the authors. For comparison with reproducibility studies in other fields reported above, the seven fully reproducible papers should be divided by the 25 reviewed for reproducibility, resulting in a 28% success rate. Expanding the denominator to include the total number of research articles published in JAS from May 2024 (when the first article to pass the reproducibility review, @herskindComputationalLinguisticMethodology2024, was published) to January 2025 (n = 97), which are articles that could have been eligible for reproducibility review had the authors used an open source programming language (e.g. instead of commercial software such as Microsoft Excel or SPSS, etc.) the success rate is 7%. In any case, the computational reproducibility of archaeological research is generally on the low end of the distribution of values obtained from a variety of hard and soft sciences.

@@

Despite the relatively small number of reproducibility reviews reported on here, there are patterns of common issues that point to a small set of simple tasks authors can do that have high potential to increase reproducibility. The problem of incomplete materials can be tackled in several basic ways. First, authors should use a clear and logical folder structure for their code and data. There are many excellent, simple, and widely-used templates that authors can choose from, e.g. @doi:10.1080/00031305.2017.1375986, @figueiredoSimpleKitUse2022, @Cookiecutter, @cooperGuideReproducibleCode2017 and @wilsonGoodEnoughPractices2017

Second, for analyses that are not highly time-consuming (which was over 90% of the sample), authors should re-run their code more than once, and ideally not on the same computer (e.g. by another co-author of the paper), before submission to confirm everything works as expected. This would ensure portability and likely also prevent issues relating to path and function errors. Complex and time-consuming analyses should use pipeline or workflow management tools, e.g. GNU Make, Luigi, Snakemake, or Targets, to document the relationship of the files and folders in a machine-readable format and simplify running and re-running code [@wrattenReproducibleScalableShareable2021; @landau2021targets].

Third, authors should include in their compendium a README document that describes to readers the folders and files contained in the project. The README file is typically the first file that a reader will look at in a compendium so it should include brief instructions to guide the user to a successful reproduction of the original results (e.g. what order to run the code files in). A README should also briefly describe the contents of the compendium, where other necessary files can be obtained (e.g. data files not included in the compendium because they require registration), and the key software packages needed and the version numbers that the authors used.

Fourth, authors should ensure clear, direct and obvious connections between their code and the results they present in their paper. One simple way to do this is to have one code file for each figure and table, and name the code files with the figure or table number and some key words in the caption. Another way some authors are accomplishing this is by using literate programming tools, such as Quarto and Jupyter notebooks [@quarto; @kluyver2016jupyter] that enable the research narrative and code for data analysis to be woven together in one document. Quarto was a popular tool among the JAS papers in the reproducibility review sample, for example, @vernonMethodDefiningDispersed2024 and @ragnoSheepGoatsTaxonomic2024 wrote their entire manuscripts using Quarto. These four simple steps have high potential to substantially increase the computational reproducibility of archaeological science.

Our finding that dependencies are a common cause of irreproducible results is consistent with previous studies that have identified this as a widespread weakness in communicating computationally intensive research [@trisovicLargescaleStudyResearch2022; @samuelComputationalReproducibilityJupyter2024]. In our sample, issues relating to dependencies are strongly associated with the use of Python. One possible reason for this is that relative to R, Python uses more package managers, more environments, and deeper dependency chains with more complex inter-dependencies that change more rapidly [@decan2016topology; @decanEmpiricalComparisonDependency2019; @korkmazModelingImpactPython2020]. Another reason may be that there is a bigger and more established community of R users in archaeology [@schmidtToolDrivenRevolutionsArchaeological2020; @batistOpenArchaeologyOpen2024] that highly values code that is easy for others to reuse and has evolved practices to effectively communicate dependencies [e.g. @bilottiPointPatternAnalysis2024; @willExploringUtilityUnretouched2025].

The simplest way for archaeologists, to improve here is to write the names and version numbers of the software and packages they used in their README file, as we see in @herskindComputationalLinguisticMethodology2024 and @monnaStudyingSerialityMaterial2024. For more complex research projects, i.e. those using five or more packages or machine learning algorithms, authors should use dependency management tools to keep track of the packages and version numbers needed to reproduce their results. This is an active area of development, and while there are many tools currently available, the most robust and widely used include renv for R [@renv, see examples in @vernonMethodDefiningDispersed2024 and @ragnoSheepGoatsTaxonomic2024] and conda and poetry for Python [@conda2023; @poetry2023].

A more comprehensive solution, and the leading best practice for managing dependencies in many computationally intensive fields using Python in particular, is to include a Dockerfile in the compendium [@moreauContainersComputationalReproducibility2023]. This is a set of a machine- and human-readable instructions that enables a user to recreate the author's computational environment (including those requirements beyond the R or Python packages) on another computer [@nustTenSimpleRules2020]. Dockerfiles are gradually being adopted by archaeologists, see @cremaModellingDiffusionInnovation2024 and @liaoApplyingMaskRCNN2024 for examples, and most of our reproducibility reviews include a recommendation that the author include a Dockerfile to manage complex dependencies efficiently.

@@

I have presented a bibliometric analysis on the status of archaeology as a science, showing that it is If archaeological science is to continue to progress through increased consensus through the accumulation of reliable facts and methods, it is essential for researchers to take computational reproducibility seriously. Computers have become a central field and laboratory instrument for so much of our work, so we have an ethical duty to document how we change our data as it flows through silicon just as carefully as we document the operating parameters of a mass spectrometer or any other field or laboratory instrument. However, the current state of quantitative archaeology, with most researchers not using open source code, is comparable to the secrecy of alchemy prior to the emergence of chemistry. Abandoning this habit of secrecy in favour of reproducibility is vital if we are to avoid a future where our journals are filled with pretty pictures that the reader has no hope of repeating or adapting in their own work. It is time for computational reproducibility to be considered a minimum requirement for evaluating the integrity and usefulness of empirical results.

Computational reproducibility is not a panacea, results that are fully reproducible can contain errors and fraud. It is no guarantee of code quality, or that statistics have been used appropriately (cf. Vaiglova this issue), or that data management is consistent with FAIR and CARE principles [cf. @carrollOperationalizingCAREFAIR2021]. It is also time-consuming for authors to ensure their computational work can be reproduced, and for reviewers to evaluate. In an professional environment where job security and career progression is often associated with pressure to publish many high-impact papers, demands for authors to spend time on reproducibility, resulting in less time for publishing more papers, may seem frustrating. This may seem especially unfair to early career researchers on short-term contracts, who may feel the goalposts for career success are being moved and that they are being asked to do work their graduate training has not prepared them for. This highlights the need for a culture shift among senior archaeological scientists to value reproducibility in hiring and promotion decision-making. Professors should contribute to this culture shift by nurturing a positive culture of reanalysis and reproducibility in their graduate teaching and evaluation [e.g. @marwickHowUseReplication2020].

@@

@@
